Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-03-19 numerical-starformation.tex

\begin{equation}
t_{cool} < t_{dyn} \equiv \sqrt{3 \pi / 32G \rho_{tot}},
\end{equation}

@@

Where $\rho_b$ is the baryon density in that cell, $\bar{\rho_b}$ is
the mean baryon density in the simulation, $\eta$ is the user-defined
overdensity threshold, $\vec{v}$ is the velocity of the gas in the
cell, $\rho_{d}$ is the dark matter density, $\rho_{total} = \rho_{b}
+ \rho_{d}$, $m_{b}$ and $m_{j}$ are the baryonic mass in the cell and
jeans mass of the cell, and C is the isothermal sound speed in the
cell.  If all of these criteria are met, the mass of a star particle
is calculated as \(m_{*} = m_{b} \frac{ \Delta t}{ t_{dyn} } f_{*eff}
\), where $f_{*eff}$ is the star formation efficiency parameter.

If $m_{*}$ is greater than a minimum star mass $m_{*min}$, a particle
is created and given several attributes: mass, a unique index number,
the time of formation $t_{form}$, the local dynamical free-fall time
$t_{dyn}$ and the metallicity fraction of the baryon gas in the cell
$f_{Zb}$.  There is a user-defined minimum dynamical time
$T_{dyn,min}$ which is observationally motivated and affects the
feedback rates (see below).  The particle is placed in the center of
the cell and given the same peculiar velocity as the gas in the cell,
and is then treated in the same manner as the dark matter particles.
An amount of baryon gas corresponding to the new particle's mass is
removed from the cell.

@@

\begin{eqnarray}  
\Delta m_{sf} =  m_{*} \left[ \left(1 + \frac{t-t_{form}}{t_{dyn}}  \right)
%\nonumber \times \\ % uncomment for 2-column
		   \exp{\left( \frac{-(t-t_{form})}{t_{dyn}} \right) }
%                    \nonumber \\ 
	       - \left( 1 + \frac{t+ \Delta t-t_{form}}{t_{dyn}} \right) 
%	       \nonumber \\ 
                    \exp{ \left( \frac{-(t+\Delta t-t_{form})}{t_{dyn}} \right)  }
                   \right]
\end{eqnarray}

@@

\begin{eqnarray}
\int_{t}^{t+Dt} \frac{dM}{dt} dt = \int_{t}^{t+Dt} m_{*} 
\frac{dt}{t_{tyn}} \left( \frac{t-t_{form}}{t_{dyn}} \right) 
%\nonumber \times \\ 
\exp{ \left( - ~ \frac{t-t_{form}}{t_{dyn}} \right) }
\end{eqnarray}

During this timestep, we assume that the star particle feeds back
metal-enriched gas and thermal energy from supernovae and from stellar
winds.  Since massive stars have very short lifetimes, we assume that
there is an immediate feedback of some fraction $f_{SN}$ of the rest
energy from the stars created that timestep into the baryon gas, such
that $E_{add} = f_{SN} \Delta m_{sf} c^2$, where c is the speed of
light.  In addition, a fraction $f_{Z*}$ of the metal from the star
particle is fed back into the baryon gas, which takes into account the
effects of metal recycling.  Finally, a fraction of the mass $f_{m*}$
from all stars (rather than just supernovae) is fed back into the gas
along with momentum in order to simulate the mass ejection from
non-exploding stars via stellar winds.

There are six user-defined parameters in this algorithm: three deal
with the star formation ($\eta$, $m_{*min}$ and $T_{dyn,min}$), and
three deal with feedback ($f_{SN}$, $f_{Z*}$ and $f_{m*}$).  Some of
these parameters are completely free, while others can be guided by
observation or theory.  For example, the supernova feedback parameter,
$f_{SN}$, can be constrained assuming that, for every $200 M_\odot$ of
stars created, one supernova occurs, and this event feeds back
approximately $10^{51}$ ergs of thermal energy, giving:

\begin{equation}
f_{SN} = \frac{10^{51} ergs}{200~M_\odot c^2} \simeq 3 \times 10^{-6}
\end{equation}

The metal yield $f_{Z*}$, defined as the mass in metals produced per
unit mass of stars created, can be constrained by, e.g., the
theoretical model of \citet{1995ApJS..101..181W}.  This model suggests
that $f_{Z*} = 0.02$ is an appropriate number.  The minimum dynamical
time is set to be $T_{dyn,min} = 10^7$ ~years to agree with timescales
seen in nearby OB associations.

The other parameters, such as the overdensity threshold $\eta$,
minimum star mass $m_{*min}$, and mass ejection fraction $f_{m*}$ are
not well constrained either theoretically or observationally.  Indeed,
$m_{*min}$ is a purely numerical parameter designed to keep the code
from producing too many star particles, and thus has no observational
or theoretical counterpart.  The $\eta$ parameter, on the other hand,
nominally has physical meaning (i.e., the density above which star
formation must occur on a very short timescale in a self-gravitating
cloud); however, in the vast majority of simulations the densities
that are reachable are nowhere near the densities of protostellar
clouds, and thus this parameter becomes a rough proxy for finding the
densest environments in a given simulation.  Given the
simulation-dependent nature of these parameters, it is critical to set
them carefully, and with some sensitivity testing.

@@

The Kravtsov method of star particle creation is designed to reproduce
the global Schmidt law of star formation~\citep{2003ApJ...590L...1K,
1959ApJ...129..243S}.  This algorithm is deliberately minimal, and is
explicitly geared towards modeling star formation in a
phenomenological way on kiloparsec scales.  Stars are assume to form
with a characteristic gas timescale $\tau_*$ such that $\dot{\rho}_* =
\rho_{gas}/\tau_*$ where $\rho_{gas}$ and $\rho_*$ are the baryon gas
and stellar densities, respectively.  This ``constant efficiency''
model on the scale of star formation regions is well motivated
observationally \citep{1996AJ....112.1903Y,2002ApJ...569..157W}.  Star
formation is only allowed to take place in very dense regions with
$\rho_{gas} \geq \rho_{SF}$, where $\rho_{SF}$ is a constant proper
(as opposed to comoving) density threshold.  No other criteria are
imposed.  Typical choices for $\tau_*$ and $\rho_{SF}$ are $\tau_* =
4$ Gyr and $\rho_{SF} = 1.64~$M$_\odot$~pc$^{-3}$~$(n_H \sim
50$~cm$^{-3})$.  The adopted timescale is derived from the
observationally-determined normalization of the Schmidt law, and the
density threshold is determined by observations of star forming
regions on $\sim 100$ pc scales.

Algorithmically, the star formation events in Kravtsov's code are
assumed to occur once every global time step $\Delta t_0 \leq 10^7$.
In cells where star formation is determined to occur (i.e. $\rho_{gas}
\geq \rho_{SF}$), star particles with a mass of $m_* = \dot{\rho}_*
V_{cell} \Delta t_0$ (where $V_{cell}$ is the volume of the mesh cell)
are assumed to form instantaneously in a manner similar to that
described in Section~\ref{sec:starform_cen}.  The \enzo\
implementation of this algorithm is similar, except that instead of
forming stars only at the root grid time step, we allow stars to form
at the time step of the highest level of resolution at any particular
point in space.  As can be seen from the equation for $m_*$ above,
this can result in very small stellar masses.  To avoid memory and
processor time issues related to having very large numbers of star
particles we impose a threshold mass $M_{*,min}$ such that a star
particle only forms if $m_* \geq M_{*,min}$.  An appropriate choice of
$M_{*,min}$ does not significantly change the star overall star
formation history of a simulation, though it may delay the onset of
star formation in a given cell relative to a simulation without a
particle mass threshold.

Each ``star particle'' is assumed to represent an ensemble of stars
and is treated as a single-age stellar population (as in the previous
section).  Kravtsov assumes that the IMF is described by a Miller \&
Scalo functional form with stellar masses between $0.1$ and
$100~M_\odot$ \citep{1979ApJS...41..513M}.  All stars in this IMF with
$M_* > 8~M_\odot$ deposit $2 \times 10^{51}$ ergs of thermal energy
and a mass $f_z M_*$ of metals into the cell in which they form
without delay, with $f_z \equiv min(0.2, 0.01~M_*-0.06)$
(i.e. instantaneous deposition of metals).  The definition of $f_z$ is
a rough approximation of the results of \citet{1995ApJS..101..181W}.

@@

% Specific example: H2-regulated
\subsubsection{H2-regulated method}
\label{sec:starform_H2reg}

@@

\begin{equation} 
m_p = \epsilon \rho_{gas} (\Delta x_{m})^3 \frac{\Delta t_0}{t_*}
\end{equation}

@@

% Specific example: Pop III
\subsubsection{Population III}
\label{sec:starform_pop3}

Unlike in the previous sections, under some circumstances it is both
possible and desirable to simulate stars individually, rather than
treating particles as ensembles.  One particular example of this is
Population III star formation
\citep{ABN02,2007ApJ...654...66O,2008ApJ...685...40W,2009Sci...325..601T},
where a given halo may only form one star of primordial composition.
To accommodate this, \citet{2007ApJ...659L..87A,2008ApJ...685...40W, 2012MNRAS.427..311W} have created an
algorithm within the \enzo\ star formation framework that forms
Population III stars directly.  Using criteria similar
to~\ref{CO1992}, a star particle forms when a cell has all of the
following criteria:

\begin{enumerate}
\item a baryon overdensity of $5 \times 10^5$ (corresponding to
  roughly $10^3$ g cm$^{-3}$ at $z=10$),
\item a converging gas flow (\nabla \cdot \vec{v} < 0), and
\item a molecular hydrogen fraction f$_{H2} > 5 \times 10^{-4}$,
\end{enumerate}

@@

\begin{equation}
f(log M) dM = M^{-1.3} \exp \Big[ -\Big( \frac{M_{char}}{M}  \Big)^{1.6} \Big]
\end{equation}

>>>>

\begin{equation}
t_{\rm cool} < t_{\rm dyn} \equiv \sqrt{3 \pi / 32G \rho_{\rm tot}},
\end{equation}

@@

Where $\rho_b$ is the baryon density in that cell, $\bar{\rho_b}$ is
the mean baryon density in the simulation, $\eta$ is the user-defined
overdensity threshold, $\vec{v}$ is the velocity of the gas in the
cell, $\rho_{d}$ is the dark matter density, $\rho_{\rm tot} =
\rho_{b} + \rho_{d}$, $m_{b}$ and $m_{j}$ are the baryonic mass in the
cell and jeans mass of the cell, and C is the isothermal sound speed
in the cell.  If all of these criteria are met, the mass of a star
particle is calculated as \(m_{*} = m_{b} \frac{ \Delta t}{ t_{\rm
    dyn} } f_{\rm *eff} \), where $f_{\rm *eff}$ is the star formation
efficiency parameter.

If $m_{*}$ is greater than a minimum star mass $m_{\rm *min}$, a particle
is created and given several attributes: mass, a unique index number,
the time of formation $t_{\rm form}$, the local dynamical free-fall time
$t_{\rm dyn}$ and the metallicity fraction of the baryon gas in the cell
$f_{\rm Zb}$.  There is a user-defined minimum dynamical time
$T_{\rm dyn,min}$ which is observationally motivated and affects the
feedback rates (see below).  The particle is placed in the center of
the cell and given the same peculiar velocity as the gas in the cell,
and is then treated in the same manner as the dark matter particles.
An amount of baryon gas corresponding to the new particle's mass is
removed from the cell.

@@

\begin{eqnarray}  
\Delta m_{\rm sf} =  m_{*} \left[ \left(1 + \frac{t-t_{\rm
        form}}{t_{\rm dyn}}  \right)
%\nonumber \times \\ % uncomment for 2-column
		   \exp{\left( \frac{-(t-t_{\rm form})}{t_{\rm dyn}} \right) }
%                    \nonumber \\ 
	       - \left( 1 + \frac{t+ \Delta t-t_{\rm form}}{t_{\rm dyn}} \right) 
%	       \nonumber \\ 
                    \exp{ \left( \frac{-(t+\Delta t-t_{\rm
                            form})}{t_{\rm dyn}} \right)  }
                   \right]
\end{eqnarray}

@@

\begin{eqnarray}
\int_{t}^{t+Dt} \frac{dM}{dt} dt = \int_{t}^{t+Dt} m_{*} 
\frac{dt}{t_{\rm dyn}} \left( \frac{t-t_{\rm form}}{t_{\rm dyn}} \right) 
%\nonumber \times \\ 
\exp{ \left( - ~ \frac{t-t_{\rm form}}{t_{\rm dyn}} \right) }
\end{eqnarray}

During this timestep, we assume that the star particle feeds back
metal-enriched gas and thermal energy from supernovae and from stellar
winds.  Since massive stars have very short lifetimes, we assume that
there is an immediate feedback of some fraction $f_{\rm SN}$ of the rest
energy from the stars created that timestep into the baryon gas, such
that $E_{\rm add} = f_{\rm SN} \Delta m_{\rm sf} c^2$, where c is the speed of
light.  In addition, a fraction $f_{Z*}$ of the metal from the star
particle is fed back into the baryon gas, which takes into account the
effects of metal recycling.  Finally, a fraction of the mass $f_{m*}$
from all stars (rather than just supernovae) is fed back into the gas
along with momentum in order to simulate the mass ejection from
non-exploding stars via stellar winds.

There are six user-defined parameters in this algorithm: three deal
with the star formation ($\eta$, $m_{\rm *min}$ and $T_{\rm
  dyn,min}$), and three deal with feedback ($f_{\rm SN}$, $f_{Z*}$ and
$f_{m*}$).  Some of these parameters are completely free, while others
can be guided by observation or theory.  For example, the supernova
feedback parameter, $f_{\rm SN}$, can be constrained assuming that, for
every $200 M_\odot$ of stars created, one supernova occurs, and this
event feeds back approximately $10^{51}$ ergs of thermal energy,
giving:

\begin{equation}
f_{\rm SN} = \frac{10^{51}\, \mathrm{erg}}{200~M_\odot\, c^2} \simeq 3 \times 10^{-6}
\end{equation}

The metal yield $f_{Z*}$, defined as the mass in metals produced per
unit mass of stars created, can be constrained by, e.g., the
theoretical model of \citet{1995ApJS..101..181W}.  This model suggests
that $f_{Z*} = 0.02$ is an appropriate number.  The minimum dynamical
time is set to be $T_{\rm dyn,min} = 10^7$ ~years to agree with timescales
seen in nearby OB associations.

The other parameters, such as the overdensity threshold $\eta$,
minimum star mass $m_{\rm *min}$, and mass ejection fraction $f_{m*}$ are
not well constrained either theoretically or observationally.  Indeed,
$m_{\rm *min}$ is a purely numerical parameter designed to keep the code
from producing too many star particles, and thus has no observational
or theoretical counterpart.  The $\eta$ parameter, on the other hand,
nominally has physical meaning (i.e., the density above which star
formation must occur on a very short timescale in a self-gravitating
cloud); however, in the vast majority of simulations the densities
that are reachable are nowhere near the densities of protostellar
clouds, and thus this parameter becomes a rough proxy for finding the
densest environments in a given simulation.  Given the
simulation-dependent nature of these parameters, it is critical to set
them carefully, and with some sensitivity testing.

@@

The Kravtsov method of star particle creation is designed to reproduce
the global Schmidt law of star formation~\citep{2003ApJ...590L...1K,
  1959ApJ...129..243S}.  This algorithm is deliberately minimal, and
is explicitly geared towards modeling star formation in a
phenomenological way on kiloparsec scales.  Stars are assume to form
with a characteristic gas timescale $\tau_*$ such that $\dot{\rho}_* =
\rho_{\rm gas}/\tau_*$ where $\rho_{\rm gas}$ and $\rho_*$ are the
baryon gas and stellar densities, respectively.  This ``constant
efficiency'' model on the scale of star formation regions is well
motivated observationally
\citep{1996AJ....112.1903Y,2002ApJ...569..157W}.  Star formation is
only allowed to take place in very dense regions with $\rho_{\rm gas}
\geq \rho_{\rm SF}$, where $\rho_{\rm SF}$ is a constant proper (as
opposed to comoving) density threshold.  No other criteria are
imposed.  Typical choices for $\tau_*$ and $\rho_{\rm SF}$ are $\tau_*
= 4$ Gyr and $\rho_{\rm SF} = 1.64~$M$_\odot$~pc$^{-3}$~$(n_{\rm H}
\sim 50$~cm$^{-3})$.  The adopted timescale is derived from the
observationally-determined normalization of the Schmidt law, and the
density threshold is determined by observations of star forming
regions on $\sim 100$ pc scales.

Algorithmically, the star formation events in Kravtsov's code are
assumed to occur once every global time step $\Delta t_0 \leq 10^7$.
In cells where star formation is determined to occur (i.e. $\rho_{\rm
  gas} \geq \rho_{\rm SF}$), star particles with a mass of $m_* =
\dot{\rho}_* V_{\rm cell} \Delta t_0$ (where $V_{\rm cell}$ is the
volume of the mesh cell) are assumed to form instantaneously in a
manner similar to that described in Section~\ref{sec:starform_cen}.
The \enzo\ implementation of this algorithm is similar, except that
instead of forming stars only at the root grid time step, we allow
stars to form at the time step of the highest level of resolution at
any particular point in space.  As can be seen from the equation for
$m_*$ above, this can result in very small stellar masses.  To avoid
memory and processor time issues related to having very large numbers
of star particles we impose a threshold mass $M_{\rm *,min}$ such that
a star particle only forms if $m_* \geq M_{\rm *,min}$.  An
appropriate choice of $M_{\rm *,min}$ does not significantly change
the star overall star formation history of a simulation, though it may
delay the onset of star formation in a given cell relative to a
simulation without a particle mass threshold.

Each ``star particle'' is assumed to represent an ensemble of stars
and is treated as a single-age stellar population (as in the previous
section).  Kravtsov assumes that the IMF is described by a Miller \&
Scalo functional form with stellar masses between $0.1$ and
$100~M_\odot$ \citep{1979ApJS...41..513M}.  All stars in this IMF with
$M_* > 8~M_\odot$ deposit $2 \times 10^{51}$ erg of thermal energy
and a mass $f_z M_*$ of metals into the cell in which they form
without delay, with $f_z \equiv \min(0.2, 0.01~M_*-0.06)$
(i.e. instantaneous deposition of metals).  The definition of $f_z$ is
a rough approximation of the results of \citet{1995ApJS..101..181W}.

@@

% Specific example: H2-regulated
\subsubsection{\HH-regulated method}
\label{sec:starform_H2reg}

@@

\begin{equation} 
m_p = \epsilon \rho_{\rm gas} (\Delta x_{m})^3 \frac{\Delta t_0}{t_*}
\end{equation}

@@

% Specific example: Pop III
\subsubsection{Population III star formation}
\label{sec:starform_pop3}

Unlike in the previous sections, under some circumstances it is both
possible and desirable to simulate stars individually, rather than
treating particles as ensembles.  One particular example of this is
Population III star formation
\citep{ABN02,2007ApJ...654...66O,2008ApJ...685...40W,2009Sci...325..601T},
where a given halo may only form one star of primordial composition.
To accommodate this, Enzo contains a star formation algorithm that
forms Population III stars directly \citep{2007ApJ...659L..87A,
  2008ApJ...685...40W, 2012MNRAS.427..311W}.  Using criteria similar
to~\citet{CO1992}, a star particle forms when a cell has all of the
following criteria:

\begin{enumerate}
\item a baryon overdensity of $5 \times 10^5$ (corresponding to
  roughly $10^3$ g cm$^{-3}$ at $z=10$),

\item a converging gas flow ($\nabla \cdot \vec{v} < 0$), and

\item a molecular hydrogen fraction f$_{\rm H2} > 5 \times 10^{-4}$,
\end{enumerate}

@@

\begin{equation}
f(\log M) dM = M^{\alpha} \exp \Big[ -\Big( \frac{M_{\rm char}}{M}
\Big)^{\beta} \Big]
\end{equation}
